---
title: "Bonanza Snowshoe Hares: exploratory report"
author: "Cristina Robinson"
date: "11/7/2021"
output: 
  html_document:
    theme: yeti
    toc: yes
---

### Overview
The snowshoe hares in Bonanza Creek Experimental Forest Long Term Ecological Research site (LTER) data is part of the US Long Term Ecological Research Network. 

The data are a fundamental component of conducting research to study the response of the boreal biome to global climate change. The Bonanza LTR program collect various types of environmental data from research locations throughout the boreal forest of Alaska with their near-real-time climate station. Data collected include temperature precipitation, wind, solar radiation, soil temperature, and soil moisture, as well as data for hydrology, animal populations, gas fluxes, tree growth, and vegetation communities. 

This report will explore the counts and sizes of juvenile snowshoe hares recorded from 1998-2012. 

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse) # install packages
library(here)
library(lubridate)
library(kableExtra)
library(effsize)
library(broom)
```


```{r}
bnzhares_data <- read_csv(here("data", "bonanza_hares.csv")) # read in data
```

### 2. Juvenile snowshoe hare initial investigation: this section looks at the total number of juvenile hare trappings during each study year.
```{r, include = FALSE}
# create columns for total number of juvenile hare trappings per year
hares_year <- bnzhares_data %>% # new column for years
  mutate(date = mdy(date)) %>% # change date format
  mutate(year = year(date))

hares_juvenile <- hares_year %>% # store new data frame of just juvenile hares
  filter(age == "j") %>%
  mutate(sex = case_when(sex == "f" ~ "Female", # mutate for graph 
                         sex == "m" ~ "Male")) %>%
  mutate(grid = case_when(grid == "bonrip" ~ "Riparian Habitat", # mutate for graph labels
                          grid == "bonmat" ~ "Mature Trees", 
                          grid == "bonbs" ~ "Black Spruce"))

hares_traps <- hares_year %>% # new column for traps
  select(year, age) %>%
  filter(age == "j") %>% # filter for juvenile hares
  group_by(year) %>% # group by year 
  summarize(number = n())

hares_traps %>%
  kbl() %>%
  kable_styling()
```

```{r}
ggplot(data = hares_traps, # plot number of hare traps per year 
       aes(x = year, y = number)) +
  geom_col(color = "gray", fill = "aquamarine") + # customize colors
  scale_x_continuous(breaks = 1999:2012) +
  theme(panel.background = element_rect(fill= "lavenderblush1", color = "purple4")) + # customize colors
  theme(plot.background = element_rect(fill = "plum3", color = "purple4")) + # customize colors
  labs(x = "Year", # add lvels
       y = "Number of Traps", 
       title = "Annual Juvenile Snowshoe Hare Traps", # add title
       caption = "Figure 1. Juvenile Snowshoe Hare Traps from 1999-2012") + # add figure caption
theme(plot.title = element_text(hjust = 0.5)) # center title
```

Then, in 3 - 4 sentences below your figure, describe major takeaways from your graph about total snowshoe hare trappings at the site. You might consider:

- The minimum count of juvenile hare trappings was 2 traps in 2010, while the maximum count of juvenile hare trappings was 126 in 1999. The mean annual number of juvenile hares trapped was 31.5, while the median number of juvenile hares trapped was 18.5 traps per year. The general trend of this data is that there are less juvenile hare traps put out as the years go in, seen by the highest number (126) of hare traps placed in 1999 and the lowest number (2) of hare traps placed in 2010. 

- These are absolute counts, but this count is also impacted by effort (e.g. the number of days and number of traps in the study each year). What would you suggest be done moving forward (you are not expected to actually do this) to standardize the juvenile hare population in future analyses? I would suggest 

### 2. Juvenile snowshoe hare initial investigation: this section looks at juvenile hare weights by sex and research site, including groups where these variables were not recorded. 

```{r}
juvehare_weights <- hares_juvenile %>% # create new data frame with just juvenile weights 
  select(weight, sex, grid) # select for weight, sex, research site (grid)
  ggplot(data = juvehare_weights, aes(x = sex, y = weight, color = sex)) + # plot data
    geom_jitter(aes(color = sex), size = 1) + # jitter plot
    geom_boxplot(fill = NA) + # boxplot
    scale_color_manual(values = c("orchid2", "darkseagreen3", "lightsteelblue4")) + # customize data colors 
    theme_minimal() + # customize background 
    facet_wrap( ~ grid) + # facet wrap
    labs (x = "Sex", # label graph
       y = "Sites", 
       title = "Juvenile Snowshoe Hare Sex and Weight per Site", # add title
       caption = "Figure 2. Sex of Juvenile Snowshoe Hares by Weight per Site") + # add caption
    
theme(plot.title = element_text(hjust = 0.5)) 
```

#### Takeaways: Then, in 2 - 3 sentences below your visualization(s) describe in text the main takeaways that you would want the audience to understand from your figures.

### 3. Juvenile snowshoe hare initial investigation: this section further looks at juvenile hare weights by sex comparing mean weights.
```{r}
juve_hare_weight_compare <- hares_juvenile %>% # new data frame 
  select(sex, weight) %>% # select sex and weight of juvenile hares
  group_by(sex) %>% # group by male and female 
  drop_na(weight) # filter out NA weights 

juve_hare_weight_summary <- juve_hare_weight_compare %>% # new data frame 
  summarize(mean_weight = mean(weight, na.rm = TRUE), # summary of statistics 
            median_weight = median(weight, na.rm = TRUE),
            sd_weight = sd(weight, na.rm = TRUE), #filter out NA weights 
            n = n()) #how many there are 

juve_hare_weight_summary %>% # create summary table 
  kbl(col.names = c("Sex", # add column names to summary table 
                    "Mean Weight (g)",
                    "Median Weight (g)",
                    "Standard Deviation",
                    "Sample Size (n)"),
      caption = "Table 1. Statistical summary of Descriptive statistics (mean, median, standard deviation, and sample size) for juvenile snowshoe hare population. Data: Kielland et al. 2017. ") %>% # add figure caption
  kable_styling(full_width = FALSE)
  
```
A 2 - 3 sentence comprehensive statement of the means comparison, including at least:

```{r, include = FALSE} 
# t-test
male_juvenile_hares <- hares_juvenile %>% # male juvenile hare data 
  filter(sex == "Male") %>%
  drop_na(weight) %>%
  pull(weight)

female_juvenile_hares <- hares_juvenile %>% # female juvenile hare data 
  filter(sex == "Female") %>% 
  drop_na(weight) %>%
  pull(weight)

t.hare <- t.test(male_juvenile_hares, female_juvenile_hares) # unpaired two sample t-test

cohen.hare <- cohen.d(male_juvenile_hares, female_juvenile_hares) # Cohen's d

t.hare # results 
cohen.hare # results 
```

```{r, include = FALSE}
# Visual data exploration with histograms and QQ-plots
ggplot(data = hares_juvenile, aes(x = weight)) +
  geom_histogram(bins = 15) + # plot histogram to view if data is normally distributed
  facet_wrap(~ sex)

ggplot(data = hares_juvenile, aes(sample = weight)) +
  geom_qq() + # create qq plot to view if data is normally distributed
  facet_wrap(~ sex)

```

### 4. Juvenile snowshoe hare initial investigation: this section looks at the relationship between hare hind food length and weight for all juvenile hares in the study. 

```{r}
ggplot(data = hares_juvenile, # plot hare weight vs. hind foot length
       aes(x = hindft, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm", # add regression
              color = "paleturquoise1", # customize color
              size = 0.5,
              fill = "lavenderblush4", # customize color
              alpha = 0.5) +
  theme_light() +
  labs (x = "Hind Foot Length (mm)", # label graph
       y = "Weight (g)", 
       title = "Juvenile Snowshoe Hare Hind Foot Length Versus Weight", # add title
       caption = "Figure 3. Weights and Hind Foot Lengths of Juvenile Snowshoe Hares") + # add caption 
theme(plot.title = element_text(hjust = 0.5)) # center title 
```

R squared = 0.299

```{r}
# Pearson's 
hares_juvenile_cor <- cor.test(hares_juvenile$weight, hares_juvenile$hindft) # Pearson's r correlation

hares_cor_tidy <- tidy(hares_juvenile_cor) # tidy version of correlation output
```

```{r, include= FALSE}
# metrics at a glance: 
hares_juve_lm <- lm(weight ~ hindft, data = hares_juvenile) #linear model of juvenile hares
hares_juve_lm_out <- broom::glance(hares_juve_lm) # r squared
hares_juve_lm_out
```

### Summary
Exploratory data analysis reveals the following initial findings:
- This population 

#### **Citations**: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14 
